### 2016-07-04 store-consistency.tex
various edits to regain space

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
off-the-shelf relational databases. Similar approach can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are relegated to the supplementary in the interest of space.}.

@@

\begin{figure}
\centering
\subcaptionbox {
  Replica model
  \label{fig:ec-theirs}
} [
  0.4\columnwidth
] {
  \includegraphics[scale=0.7]{Figures/ec-theirs}

}
\hspace*{0.1in}
\subcaptionbox {
  Subset model
  \label{fig:ec-ours}
}{
  \includegraphics[scale=0.45]{Figures/ec-ours}
} \caption{In the replica model, operation $\op$ generates effect
$\eta$ at replica $R_1$, which is then merged to $R_2$. If the
\emph{store is {\sc cc}}, then $R_2$'s state at merge event is same or
larger than $R_1$'s state at generation event (the difference is
highlighted). In our subset model, $\op$ witnesses $S_1 \subseteq
\E.\A$ and generates $\eta$, which is immediately added to $\E.\A$. A
later operation may witness $S_2 \subseteq \E.\A$, and if the
\emph{operation is} {\sc cc} and $\eta \in S_2$, then it also
witnesses $S_1$ (i.e., $S_1 \subseteq S_2$). } 
% Moreover, Like $R_2 - R_1$, if all effects in $S_2 - S_1$ are
% concurrent with $\eta$, i.e., $\not\exists\eta'.~\eta' \in S_2 - S_1
% \conj % visZ(\eta,\eta')$, then any pre-condition $P$ that is valid
% when $\op$ executed is also valid when $\eta$ is witnessed because
% of the stability condition.
\label{fig:ec-theirs-vs-ours}
\end{figure}

@@

$\begin{decoration}
  \hspace*{0.3in}Q : \{\committed(\C{Wd1}) \conj \committed(\C{Wd2}) \conj
                          \C{C} = \C{k-a1-a2}\}
\end{decoration}$
\end{txnimpcode}

\caption{\C{Wd1} transaction decorated with assertions}
\label{fig:wd1-decorated}
\end{figure}

@@

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. We consider the
triple associated with the assignment to \C{C} for illustration. The
pre-condition $\phi_4$ asserts different values for \C{C} and \C{v3}
(temporary alias of RHS expression) depending on whether or not
\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
this is sufficient to prove the post-condition $\phi_5$, which appears
to make a stronger claim about the value of \C{C}. The proof follows
from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
to easily conclude the following about the execution state after the
assignment:

>>>>

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
off-the-shelf relational databases. Similar approaches can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are relegated to the supplementary in the interest of space.}.

@@

\begin{figure}
\centering
\subcaptionbox {
  Replica model
  \label{fig:ec-theirs}
} [
  0.4\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/ec-theirs}

}
\hspace*{0.1in}
\subcaptionbox {
  Subset model
  \label{fig:ec-ours}
}{
  \includegraphics[scale=0.35]{Figures/ec-ours}
} \caption{\small In the replica model, operation $\op$ generates effect
$\eta$ at replica $R_1$, which is then merged to $R_2$. If the
\emph{store is {\sc cc}}, then $R_2$'s state at merge event is same or
larger than $R_1$'s state at generation event (the difference is
highlighted). In our subset model, $\op$ witnesses $S_1 \subseteq
\E.\A$ and generates $\eta$, which is immediately added to $\E.\A$. A
later operation may witness $S_2 \subseteq \E.\A$, and if the
\emph{operation is} {\sc cc} and $\eta \in S_2$, then it also
witnesses $S_1$ (i.e., $S_1 \subseteq S_2$). } 
% Moreover, Like $R_2 - R_1$, if all effects in $S_2 - S_1$ are
% concurrent with $\eta$, i.e., $\not\exists\eta'.~\eta' \in S_2 - S_1
% \conj % visZ(\eta,\eta')$, then any pre-condition $P$ that is valid
% when $\op$ executed is also valid when $\eta$ is witnessed because
% of the stability condition.
\label{fig:ec-theirs-vs-ours}
\end{figure}

@@

$\begin{decoration}
  \hspace*{0.3in}Q : \{\committed(\C{Wd1}) \conj \committed(\C{Wd2}) \conj
                          \C{C} = \C{k-a1-a2}\}
\end{decoration}$
\end{txnimpcode}
\vspace*{-8pt}
\caption{\small \C{Wd1} transaction decorated with assertions}
\label{fig:wd1-decorated}
\vspace*{-12pt}
\end{figure}

@@

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. We consider the
triple associated with the assignment to \C{C} for illustration. The
pre-condition $\phi_4$ asserts different values for \C{C} and \C{v3}
(temporary alias of RHS expression) depending on whether or not
\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
this is sufficient to prove the post-condition $\phi_5$, which appears
to make a stronger claim about the value of \C{C}. The proof follows
from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
to easily conclude the following about the execution state after the
assignment:\vspace*{-12pt}
